\section{Literature Review}

The literature relevant to this study sits at the intersection of Islamic finance, portfolio optimisation and machine learning-based financial prediction. 
From an Islamic finance perspective, investment decisions are not evaluated only through return maximisation, but also through their consistency with Shariah principles and broader ethical objectives. 
The Islamic Moral Economy framework argues that Islamic finance should move beyond formal legal compliance and contribute to social welfare, justice and human-centred development \citep{asutay2007conceptualisation, asutay2012conceptualising}. 
This is particularly important in portfolio construction, where the investment universe is first restricted through Shariah screening and then optimised according to risk-return objectives. 
In this regard, Shariah-compliant portfolio optimisation can be understood as a filtered investment process in which financial efficiency is pursued within ethical boundaries. 
While earlier studies on Islamic finance have emphasised the principles of risk sharing, prohibition of riba and asset-backed transactions \citep{elgamal2006islamic, ayub2009understanding, chong2009islamic}, 
recent developments in fintech and computational finance create new opportunities to operationalise these principles through data-driven investment models. 
Therefore, the use of machine learning and deep learning in Shariah-compliant portfolio management provides a practical bridge between the ethical foundations of Islamic finance and the technical requirements of modern financial markets.


\subsection{Islamic Moral Economy and Shariah-compliant investment}

Islamic finance is not reducible to the technical prohibition of riba. Its normative
foundation rests on the concept of the Islamic Moral Economy, which argues that
financial institutions and investment practices should serve social welfare, ethical
justice and human-centred development rather than profit maximisation alone
\citep{asutay2007conceptualisation, asutay2012conceptualising}. From this
perspective, Shariah compliance is a necessary but not sufficient condition for
Islamic investment: a portfolio may satisfy formal screening criteria while failing
to realise the broader objectives of the Islamic financial system
\citep{asutay2012conceptualising}. This tension is central to the growing body of
work that evaluates whether Islamic financial institutions operate differently from
their conventional counterparts in practice \citep{chong2009islamic,
gheeraert2014islamic}.

The foundational principles governing Islamic investment include the prohibition
of riba (interest), gharar (excessive uncertainty) and maysir (speculation), as
well as the requirement for asset-backed transactions and risk sharing between
parties \citep{elgamal2006islamic, ayub2009understanding, iqbal2011islamic}.
These principles restrict the investable universe through Shariah screening, which
typically excludes firms with significant revenues from prohibited sectors such as
alcohol, tobacco, weapons and conventional financial services, as well as firms
whose debt or liquid asset ratios exceed specified thresholds
\citep{hassan2005impact}. Shariah screening therefore creates a constrained
investment universe that is smaller than the full market, and portfolio construction
must be conducted within this ethical boundary.

Evidence on the performance of Shariah-screened portfolios is mixed. Some
studies find that Islamic equity indices generate comparable or superior risk-adjusted
returns relative to conventional benchmarks over medium to long horizons,
particularly during periods of financial stress when the exclusion of highly
leveraged firms provides a degree of downside protection
\citep{hassan2005impact}. Others argue that sector and liquidity constraints
associated with Shariah screening may limit diversification and increase concentration
risk. This performance debate motivates the use of quantitative optimisation
methods that can extract diversification benefits within the screened universe,
which is one of the objectives of the present paper.

\subsection{Portfolio optimisation: from mean-variance to sparse methods}

The theoretical foundation for portfolio optimisation is the mean-variance
framework introduced by \citet{markowitz1952portfolio}. The central insight is
that diversification across imperfectly correlated assets can reduce portfolio risk
for a given level of expected return. The efficient frontier represents the set of
portfolios that are optimal in this sense, and rational investors should choose
portfolios on the frontier according to their risk preferences. This framework
has been applied across conventional and Islamic finance contexts, since the
mathematical structure of the optimisation is independent of the ethical
composition of the investable universe.

Despite its theoretical appeal, the mean-variance framework is sensitive to
estimation error. Small changes in expected return or covariance estimates can
produce large changes in optimal portfolio weights, leading to concentrated and
unstable allocations \citep{michaud1989markowitz}. This problem is especially
acute in large equity universes such as the BIST XKTUM, where the covariance
matrix must be estimated from historical data across a large number of assets.
Several responses to this challenge have been proposed in the literature,
including shrinkage estimation of the covariance matrix
\citep{ledoit2004honey}, the use of equal weights as a robust alternative
\citep{taljaard2020equal, sen2023equal}, and the application of regularisation
methods that penalise portfolio complexity and encourage sparse asset selection.

Regularisation approaches have gained particular attention in the computational
finance literature. \citet{tibshirani1996regression} introduced the LASSO
method, which applies an L1 penalty to regression coefficients to encourage
sparsity. DeMiguel et al. applied this principle to portfolio optimisation, showing
that L1-regularised portfolios can reduce the number of active positions and
limit estimation error without sacrificing out-of-sample performance. Sparse
portfolio construction is especially relevant in Shariah-compliant investing,
where the universe may be large but liquidity varies substantially across assets,
making concentrated and parsimonious allocation strategies practically important.

\subsection{Machine learning for stock price prediction}

The application of machine learning to financial prediction has grown substantially
since the early 2000s. Recurrent neural networks were among the first architectures
to be applied to financial time series, motivated by their ability to model
sequential dependencies in data where historical patterns influence future values.
The Long Short-Term Memory (LSTM) network, introduced by
\citet{hochreiter1997long}, addresses the vanishing gradient problem that limited
earlier recurrent architectures by introducing gating mechanisms that regulate
information flow across long sequences. LSTM has been applied to stock price
prediction across multiple markets and has consistently demonstrated competitive
performance relative to simpler time-series models
\citep{zhang2023stock, lu2023lstm}.

Bidirectional LSTM (Bi-LSTM) extends the LSTM architecture by processing
input sequences in both forward and backward directions and concatenating the
resulting hidden states at each time step \citep{schuster1997bidirectional}.
This allows the model to draw on both historical and future context within
the training window, which may be advantageous when the informational
content of the sequence is not strictly unidirectional. Bi-LSTM has been applied
to financial forecasting tasks including stock return prediction and cryptocurrency
price modelling, and evidence suggests that bidirectional processing can improve
accuracy relative to standard LSTM in some settings
\citep{murugesan2022bilstm, hamayel2021bilstm}.

The Gated Recurrent Unit (GRU) was proposed by \citet{cho2014learning} as
a computationally simpler alternative to LSTM. GRU merges the cell state
and hidden state into a single vector and uses update and reset gates rather
than the three-gate structure of LSTM. This reduces the number of parameters
and typically shortens training time without substantially sacrificing the
ability to model long-range dependencies. Comparative studies have found
that GRU and LSTM produce similar prediction accuracy across a range of
financial time-series tasks, with GRU offering an efficiency advantage in
resource-constrained settings \citep{hamayel2021bilstm}.

Non-sequential machine learning models have also been widely applied to stock
price prediction. \citet{breiman2001random} introduced Random Forest as an
ensemble learning method that trains multiple decision trees on random subsets
of features, producing predictions through majority voting or averaging. Random
Forest is robust to overfitting and performs well with high-dimensional inputs,
but it does not model temporal dependencies explicitly and requires that the
sequential structure of financial data be captured through feature engineering.
Support Vector Machines (SVM), introduced by \citet{cortes1995support},
identify an optimal hyperplane in the feature space and can handle non-linear
relationships through kernel functions. SVM has been applied to financial
prediction tasks including stock price classification and regression
\citep{ji2022svm}, but it can be sensitive to hyperparameter choices and
computationally demanding on large datasets.

Extreme Gradient Boosting (XGBoost) was introduced by \citet{chen2016xgboost}
as a scalable and regularised gradient boosting algorithm that builds an ensemble
of decision trees sequentially, with each tree correcting the errors of its
predecessor. XGBoost incorporates built-in L1 and L2 regularisation, handles
missing values natively and benefits from GPU-accelerated computation. It has
achieved strong results across financial prediction benchmarks, including stock
price regression and fraud detection \citep{wu2023xgboost, zhu2023lstmxgboost}.
A key advantage relative to recurrent models is that XGBoost processes engineered
feature vectors rather than sequential inputs, which makes it well suited to
datasets where informative features have been derived from the raw price series
through feature engineering.

Comparative studies that evaluate multiple model types on the same financial
dataset are particularly informative for model selection. Evidence from such
studies generally suggests that no single model dominates across all datasets
and evaluation criteria \citep{zhang2023stock, kuang2023comparison}. Recurrent
models tend to perform well when temporal dependencies are strong and the
prediction horizon is medium to long term. Tree-based ensemble models tend
to perform well when the dataset contains informative engineered features and
non-linear interactions among predictors. This observation directly motivates
the six-model comparison conducted in this paper.

\subsection{Machine learning in portfolio optimisation}

Bridging prediction and portfolio construction is more complex than it might
appear, because a model with strong forecasting accuracy does not necessarily
generate a portfolio with superior risk-adjusted performance. The relationship
between prediction error and portfolio performance depends on the distribution
of errors across assets, the covariance structure of the portfolio, and how
prediction errors compound over time \citep{brandt2010portfolio}. Several
papers have attempted to directly integrate machine learning predictions into
mean-variance optimisation, with mixed evidence on whether improved forecasting
accuracy translates into improved portfolio outcomes.

\citet{lu2023lstm} integrates LSTM-based return forecasts into portfolio
allocation and demonstrates that LSTM predictions can support cumulative
portfolio growth in Chinese equity markets. Studies using Bi-LSTM and GRU
for portfolio management have reported similar findings, although the relative
advantage over simpler baselines varies with the market and sample period
\citep{hamayel2021bilstm}. The present paper contributes to this literature
by evaluating all three sequence-based architectures within the same
Shariah-compliant equity universe and under identical optimisation conditions,
which enables a direct and controlled comparison of their portfolio-level usefulness.

A related strand of literature examines sparse portfolio construction, where
the objective is to identify a small subset of assets that captures the risk-return
efficiency of the broader universe. DeMiguel et al. show that L1-penalised
portfolios can outperform the 1/N equal-weight strategy and the unconstrained
mean-variance portfolio in out-of-sample testing when the estimation period
is limited relative to the number of assets. In large Islamic equity universes,
where investors may face constraints on the number of assets they can monitor
or trade, sparse optimisation methods are practically relevant and have received
limited direct attention in the Islamic finance literature.

\subsection{Islamic fintech and computational investment management}

The intersection of fintech and Islamic finance has attracted growing scholarly
attention, particularly in the context of financial inclusion, ethical digital
payment systems and Shariah-compliant crowdfunding platforms
\citep{rabbani2021islamic, hasan2020fintech, unal2022fintech}. However,
a majority of this literature remains conceptual or survey-based, and the
application of advanced computational methods to Islamic investment
management has received limited empirical treatment.

\citet{unal2022fintech} provide a retrospective survey of fintech development
in Islamic finance and argue that digitisation and machine learning create new
opportunities to operationalise Shariah-compliant investment principles, but
that empirical applications remain rare. \citet{hasan2020fintech} similarly
identify the lack of computational investment tools as a gap in the Islamic
fintech agenda. Recent work by \citet{rabbani2021islamic} demonstrates that
Islamic fintech platforms have begun to move beyond payments and into
investment advisory and portfolio management services, but the underlying
analytical models are rarely disclosed or evaluated in the academic literature.

The Turkish market provides a particularly relevant setting for this analysis.
Turkey has a large Muslim-majority population, an established participation
banking sector and a dedicated Islamic equity index in the BIST XKTUM.
The Islamic fintech ecosystem in Turkey has expanded in recent years, supported
by regulatory initiatives and the growth of digital participation finance platforms
\citep{sakarya2021fintech}. Despite this, the academic literature on Turkish
Islamic investment management and computational portfolio optimisation using
BIST XKTUM data remains limited, which further motivates the empirical
contribution of this paper.

\subsection{Research gap and contribution}

The review identifies three specific gaps that this paper addresses. First, while
machine learning models have been applied to stock prediction in conventional
markets, their comparative evaluation within a Shariah-compliant equity universe
using large-scale Turkish data is largely absent from the literature. Second,
existing studies that integrate ML predictions into portfolio optimisation rarely
apply L1 regularisation for sparse asset selection or evaluate portfolio outcomes
using Sharpe ratio and annualised volatility alongside terminal equity. Third,
the Islamic fintech literature has called for empirical evidence on how computational
investment tools can be applied within Shariah-compliant constraints, but few
papers deliver such evidence at the scale of the BIST XKTUM universe.

This paper addresses these gaps by providing a systematic, hyperparameter-tuned
comparison of six machine learning and deep learning models for stock price
prediction, followed by a portfolio optimisation stage that uses the predicted
returns within a sparse mean-variance framework applied to a Shariah-screened
Turkish equity universe. The results contribute to the Islamic fintech literature
by showing how prediction accuracy and portfolio performance are related but
distinct criteria, and by identifying which model types are most useful for
each purpose in the Shariah-compliant setting.